---
title: "DEMG609: Homework 1"
author: "Nick Graetz"
date: "September 13, 2017"
output: pdf_document
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Load knitr package and settings
library(knitr)
library(data.table)
library(ggplot2)
options(scipen=999)
opts_chunk$set(fig.align='center', tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=40), warning=FALSE,message=FALSE)
```

\large
1.(a) 
\vspace{12pt}

\small
```{r, echo=TRUE, eval=TRUE}
# Because the instantaneous growth rate is constant over the interval, it can be calculated as the mean annualized growth rate.
N_t1 <- 281421906
N_t2 <- 308745538
t1 <- 2000.25
t2 <- 2010.25
T <- t2 - t1
r_star <- log(N_t2 / N_t1) / T
answer_1a <- r_star
```
\large

**Answer:** `r answer_1a`

\vspace{12pt}
1.(b) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
tripling_time <- log(3) / r_star
answer_1b <- tripling_time
```
\large

**Answer:** `r answer_1b`

\vspace{12pt}
1.(c) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
# Given our growth rate is constant...
py <- ((N_t2 - N_t1) * T) / log(N_t2 / N_t1)
answer_1c <- py
```
\large

**Answer:** $`r answer_1c`$

\vspace{12pt}
1.(d) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
## What was the population size in the middle of the intercensual period?
## Given that growth rate is constant...
pop_at_t <- function(t) {
  new_pop <- N_t1 * exp(r_star * t)
  return(new_pop)
}
N_mid <- pop_at_t(T/2)
answer_1d <- N_mid
```
\large

**Answer:** $`r answer_1d`$

\vspace{12pt}
1.(e) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
## Given only midpoint population, we have to assume linear growth
py_linear <- N_mid * T

growth <- data.table(T = seq(0,1,0.1),
                     Pop = c(rep(N_mid, 11),
                             pop_at_t(0:10)),
                     Growth = c(rep('Linear', 11),
                                rep('Exp', 11)))

ggplot(data = growth) +
  geom_line(aes(x = T,
                y = Pop,
                color = Growth),
            size = 2) +
  geom_point(x = 0.5, y = N_mid, color = 'black', size = 3) + 
  ggtitle("Exp = constant growth rate given N(0) and N(T)\nLinear = given only midpoint population") + 
  theme_classic()
```
\large

**Answer:**  
Linear PY: $`r py_linear`$  
Exp PY: $`r py`$  
Difference: $`r py_linear - py`$  

\vspace{12pt}
2.(a)
\linebreak

**Answer:** $`r r_star`$  

\vspace{12pt}
3.(a) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
## Instead of 2000.25 to 2010.25, it's now 2000.25 - 2010.08333
N_t1 <- 281421906
N_t2 <- 308745538
t1 <- 2000.25
t2 <- 2010 + (1/12)
T <- t2 - t1
r_star <- log(N_t2 / N_t1) / T
answer_3a <- r_star
```
\large

**Answer:** $`r answer_3a`$

\vspace{12pt}
3.(b) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
pop_at_t <- function(t, r_star, N_t1) {
  new_pop <- N_t1 * exp(r_star * t)
  return(new_pop)
}
N_mid <- pop_at_t(t = (T/2), r_star = r_star, N_t1 = N_t1)
answer_3b <- N_mid
## Bigger, because it has to get to the same end population faster
```
\large

**Answer:** $`r answer_3b`$  
Answer to 1(d): $`r answer_1d`$  

\vspace{12pt}
3.(c) 
\linebreak

Given that $N_0$ and $N_t$ and the assumption of constant $r$ between $N_0$ and $N_t$: $N_t = N_0 * e^{rt}$ and $r = ln\biggr(\frac{N_t}{N_0}\biggl)$
\vspace{12pt}

\huge

$$N_t = N_0e^{\frac{ln\biggl(\frac{N_t}{N_0}\biggr)}{T}T}$$
\vspace{12pt}
$$N_t = N_0e^{ln\biggl(\frac{N_t}{N_0}\biggr)}$$
\vspace{12pt}
$$N_t = N_0 * \frac{N_t}{N_0}$$
\vspace{12pt}
$$N_t = N_t$$
\vspace{12pt}
$$1 = 1$$
\vspace{12pt}

\large

4.(a) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
## Reset to initial pops and dates
N_t1 <- 281421906
N_t2 <- 308745538
t1 <- 2000.25
t2 <- 2010.25
T <- t2 - t1
## Inputs:
##    j = number of times to compound per year
##    year_interval = years in the period
##    start_pop
##    end_pop
## Outputs:
##    r_j = annual geometric growth rate
calc_geo_r <- function(j, year_interval, start_pop, end_pop) {
  r_j <- j * ((end_pop / start_pop)^(1/(j*year_interval)) - 1)
  return(r_j)
}

## Apply function over j = 1, 12, 365 (annually, monthly, daily, continuously)
geom_rates <- lapply(c(1,12,365,1000000),
                     calc_geo_r,
                     year_interval = T,
                     start_pop = N_t1,
                     end_pop = N_t2)
```
\large

**Answer:**  
Annually: $`r geom_rates[[1]]`$  
Monthly:  $`r geom_rates[[2]]`$  
Daily:    $`r geom_rates[[3]]`$  
Cont:     $`r geom_rates[[4]]`$  

\vspace{12pt}
4.(b) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
calc_geo_r_2 <- function(j, annual_rate) {
  r_j <- j * ((1 + annual_rate)^(1/j) - 1)
  return(r_j)
}
r_j_cont <- calc_geo_r_2(j = 1000000, annual_rate = geom_rates[[1]])
answer_4b <- r_j_cont
```
\large

**Answer:** $`r answer_4b`$ 

\vspace{12pt}
4.(B) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
r_star_ukraine <- -0.49 / 100
halving_time <- log(0.5) / r_star_ukraine
answer_4B <- halving_time
```
\large

**Answer:** $`r answer_4B`$ 

\vspace{12pt}
4.(C) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
t1_urban <- 100
t1_rural <- 1000
r_star_urban <- 0.5
r_star_rural <- 0.05
growth_comparison <- data.table(T = seq(0,1,0.1),
                                 Pop = c(pop_at_t(0:10, r_star = r_star_urban, N_t1 = t1_urban),
                                         pop_at_t(0:10, r_star = r_star_rural, N_t1 = t1_rural)),
                                 Growth = c(rep('Urban', 11),
                                            rep('Rural', 11)))
ggplot() +
  geom_line(data = growth_comparison,
            aes(x = T,
                y = Pop,
                color = Growth),
            size = 2) +
  theme_classic()

growth_comparison <- dcast(growth_comparison, T ~ Growth, value.var = "Pop")
growth_comparison[, pop_ratio := Urban / Rural]
t1_ratio <- growth_comparison[T == 0, pop_ratio]
t2_ratio <- growth_comparison[T == 1, pop_ratio]
ratio_growth_rate <- log(t2_ratio / t1_ratio) / 10
ratio_growth_rate == (r_star_urban - r_star_rural)
```
\large

**Answer:**  
Urban growth rate: $`r r_star_urban`$  
Rural growth rate: $`r r_star_rural`$  
Difference in growth rates: $`r r_star_urban - r_star_rural`$  
Growth rate of urban/rural population ratio: $`r ratio_growth_rate`$  

\large

**Analytic Answer:**  

\huge

$$R = \frac{N_u}{N_r}$$
\large

Prove that: $Growth_{R} = Growth_u - Growth_r$

\huge

$$\frac{ln\biggl(\frac{R_t}{R_0}\biggl)}{T} = Growth_u - Growth_r$$
\vspace{12pt}
$$\frac{ln\biggl(\frac{\frac{N_{t,u}}{N_{t,r}}}{\frac{N_{0,u}}{N_{0,r}}}\biggl)}{T} = \frac{ln\biggl(\frac{N_{t,u}}{N_{0,u}}\biggl)}{T} - \frac{ln\biggl(\frac{N_{t,r}}{N_{0,r}}\biggl)}{T}$$
\vspace{12pt}
$$ln\biggl(\frac{\frac{N_{t,u}}{N_{t,r}}}{\frac{N_{0,u}}{N_{0,r}}}\biggl) = ln\biggl(\frac{N_{t,u}}{N_{0,u}}\biggl) - ln\biggl(\frac{N_{t,r}}{N_{0,r}}\biggl)$$
\vspace{12pt}

\large

Use this law to reduce all terms: $ln\biggl(\frac{a}{b}\biggr) = ln(a) - ln(b)$

\huge

$$ln\biggl(\frac{N_{t,u}}{N_{t,r}}\biggl) - ln\biggl(\frac{N_{0,u}}{N_{0,r}}\biggl)= ln\biggl(\frac{N_{t,u}}{N_{0,u}}\biggl) - ln\biggl(\frac{N_{t,r}}{N_{0,r}}\biggl)$$

\large

$$ln(N_{t,u})-ln(N_{t,r})-ln(N_{0,u})-ln(N_{0,r}) = ln(N_{t,u})-ln(N_{0,u})-ln(N_{t,r})-ln(N_{0,r})$$

\huge

$$1 = 1$$