---
title: "DEMG609: Problem Set 2"
author: "Nick Graetz"
date: "September 20, 2017"
output: pdf_document
---
  
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Load knitr package and settings
library(knitr)
library(data.table)
library(ggplot2)
library(formatR)
options(scipen=999)
#opts_chunk$set(fig.align='center', tidy=TRUE, tidy.opts=list(blank=TRUE, width.cutoff=40), warning=FALSE,message=FALSE)
#opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

\large
A.(1) 
\vspace{12pt}

\small
```{r, echo=TRUE, eval=TRUE}
# Load data.
data <- fread("C:/Users/ngraetz/Documents/repos/demg609/hw2_data.csv")
data[, pop := as.numeric(pop)]
data[, deaths := as.numeric(deaths)]

# Calculate CDR.
cdr_usa <- sum(data[country == "USA", deaths]) / sum(data[country == "USA", pop])
cdr_russia <- sum(data[country == "Russia", deaths]) / sum(data[country == "Russia", pop])
```
\large

**Answer:**  
USA: `r cdr_usa`  
Russia: `r cdr_russia`  

\vspace{12pt}
A.(2) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
data[, asdr := deaths / pop]

ggplot(data = data) +
  geom_line(aes(x = age_start,
                y = log(asdr),
                color = country),
            size = 1) +
  theme_classic()
```
\large

\vspace{12pt}
A.(3) 
\linebreak
We have to make a linear assumption as we only have midpoint populations.
\linebreak

\vspace{12pt}
A.(4) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
data[country == "USA", age_prop := pop / sum(pop)]
data[country == "Russia", age_prop := pop / sum(pop)]
ggplot(data = data) +
  geom_line(aes(x = age_start,
                y = age_prop,
                color = country),
            size = 1) +
  theme_classic()
```
\large

\vspace{12pt}
A.(5) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
cdr_usa_wt <- weighted.mean(data[country == "USA", asdr],
                            data[country == "USA", age_prop])
cdr_russia_wt <- weighted.mean(data[country == "Russia", asdr],
                               data[country == "Russia", age_prop])
cdr_usa_wt
cdr_russia_wt
```
\large

\vspace{12pt}
A.(6) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
usa_ascdr <- weighted.mean(data[country == "USA", asdr],
                           data[country == "Russia", age_prop])
```
\large

**Answer:**  
USA ASCDR: `r usa_ascdr`  
The difference must be attributable to differences in the age-specific mortality rates between the US and Russia, as we have controlled for differences in the age-specific populations.

\vspace{12pt}
A.(7) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
# Calculate CMR.
cmr <- sum(data[country == "Russia", deaths]) / sum(data[country == "USA", asdr] * data[country == "Russia", pop])
cmr
# Compare to ratio of Russia CDR and USA ASCDR.
cdr_russia / usa_ascdr
```
\large

\vspace{12pt}
A.(8) 
\linebreak

\small
```{r, echo=TRUE, eval=TRUE}
c_age <- sum((data[country == "Russia", age_prop] - data[country == "USA", age_prop]) * 
            ((data[country == "USA", asdr] + data[country == "Russia", asdr]) / 2)) 
c_asdr <- sum((data[country == "Russia", asdr] - data[country == "USA", asdr]) * 
             ((data[country == "USA", age_prop] + data[country == "Russia", age_prop]) / 2)) 
# Calculate relative contributions to the net absolute difference.
total_diff <- abs(c_age) + abs(c_asdr)
abs(c_age)/total_diff
abs(c_asdr)/total_diff
```
